---
title: "Bioinformatics prac solutions"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
theme: cosmo
---

Source: https://github.com/markziemann/SLE712_files/blob/master/solutions.Rmd

## Prac 1

### Question 1-1

Calculate the sum of all integers numbers between 500 and 600.

```{r,q1-1}

sum(500:600)

```

### Question 1-2

Calculate the sum of all the square roots of all integers between 900 and 1000

```{r,q1-2}

sum(sqrt(900:1000))

```

### Question 1-3

Create the following datasets and scatterplot a boxplot

  1. A Sample 10000 datapoints from a normal distribution with mean of 50 and SD of 5
  
  2. A Sample 10000 datapoints from a normal distribution with mean of 50 and SD of 10 

```{r,q1-3}

a <- rnorm(n=1000,mean = 50, sd = 5)

b <- rnorm(n = 1000,mean = 50, sd = 10)

boxplot(a,b,names=c("a","b"))

```


### Question 1-4

Plot a and b above as a scatterplot, and plot the trend line.

```{r,q1-4}

mylm <- lm(b~a)
plot(a,b)
abline(mylm,col="red")

```

### Qustion 1-5

Plot a and b above as histograms on the same chart.

```{r,q1-5}

par(mfrow=c(2,1))
hist(a)
hist(b)

```

## Prac 2 

### Question 2-1

Create a scatterplot of mtcars weight (x axis) versus mpg (y axis). 
Include x and y axis labels and a main heading.

```{r,q2-1,fig.height=8,fig.width=8}

plot(mtcars$wt,mtcars$mpg, xlab="weight (lb/1000)", ylab="miles per gallon", main="relationship between weight and milage")

```

### Question 2-2

Sort mtcars by weight (wt) and create a horizontal barplot of wt values so heaviest ones at shown at the top of the bar plot.
The plot should be labeled so it is clear to see which bar belong to which car.
Include an axis label and main title.

```{r,q2-2,fig.height=8,fig.width=8}

mycars <- mtcars[order(mtcars$wt),]
par(mar=c(5,12,5,3))
barplot(mycars$wt, xlim= c(0,6), horiz=TRUE,las=1,names.arg=rownames(mycars),cex.names=0.6,main="mtcars weights",xlab="car weight (lb/1000)")
grid()

``` 

### Question 2-3

Create a box plot of iris petal lengths.
Each species should be a different category.
Chart needs axis labels and a main title.

```{r,q2-3,fig.height=8,fig.width=8}

setosa <- subset(iris,Species=="setosa")
virginica <- subset(iris,Species=="virginica")
versicolor <- subset(iris,Species=="versicolor")
boxplot(list(setosa$Petal.Length,virginica$Petal.Length,versicolor$Petal.Length), names=c("Setosa","Virginica","Versicolor"),ylab="cm",main="Petal length")

```

### Question 2-4

For the TSV file located here: https://raw.githubusercontent.com/markziemann/SLE712_files/master/bioinfo_asst3_part1_files/mydata.tsv do the following:

1. Read it in and show the first 6 rows of data.

The trick here is to set `header=TRUE` and `row.names=1` so the colnames and rownames are set properly.

```{r,q2-4-1}

URL="https://raw.githubusercontent.com/markziemann/SLE712_files/master/bioinfo_asst3_part1_files/mydata.tsv"
x <- read.table(URL,header=TRUE,row.names = 1)
head(x)

```

2. Calculate the column and row means.

```{r,q2-4-2}

colMeans(x)
rowMeans(x)

```


3. Use the `cor()` command to find the correlation coefficients between the 3 data sets

ERR1720606 is most similar to ERR1720605, while ERR1720604 is less similar to the others.

```{r,q2-4-3,fig.height=8,fig.width=8}

cor(x)

library("gplots")
heatmap.2(cor(x),trace="none",scale="none",margins = c(15,15))

```

4. Make a pairs plot of the three datasets.

```{r,q2-4-4,fig.height=8,fig.width=8}

pairs(x)

```

## Session information

For reproducibility.

```{r,sessioninfo}

sessionInfo()

```