---
title: "Prac 3: Functions, conditionals, Loops and Vectorisation"
author: "Mark Ziemann"
date: "20/04/2020"
output:
    rmarkdown::html_document:
        toc: true
theme: cosmo
---
Source: https://github.com/markziemann/SLE712_files/blob/master/apply.Rmd
# Introduction
In this lesson we'll be taking your R skills to the next level with functions, conditionals, loops and vectorisation.
# Functions
A function is a stored command which can be rerun with different inputs. Functions are normally more than one line. Your function can specify many inputs. `return` is used to select what data should be returned to the user.
```{r}
cubed <- function(x) {
    x^3
}

cubed(42)
cubed(c(1,2,4))

# see how you can use functions inside other functions
cubedsum <- function(x,y) {
    z <- cubed(x) + cubed(y)
    return(z)
}
cubedsum(3,4)
```
# Conditionals
Commonly your workflow will need to adapt to different inputs. For example if a file already exists, you may not want to over-write it. This is the general format of an `if` statement in R:
```{r,eval=FALSE}
if (condition) {command to run if TRUE} 
```
Most of the time the command to run is written on a new line. Many times we need to run more than one command.
```{r,eval=FALSE}
if (condition) {
    command1 to run if TRUE
    command2 to run if TRUE
} 
```
If we only want to execute the command if FALSE, then we can add `!` to toggle the conditional.
```{r}
2>3
!2>3
```
Here is how it looks in an actual `if` statement.
```{r,eval=FALSE}
if (!condition) {
    command1 to run if FALSE
    command2 to run if FALSE
} 
```
We can add `else` to select from two options
```{r,eval=FALSE}
if (condition) {
    command1 to run if TRUE
} else {
    command2 to run if FALSE
}
```
Here is a real life example.
```{r}
if (!file.exists("mtcars.tsv")) {
  write.table(mtcars,file="mtcars.tsv",quote=FALSE,sep="\t")
  message("File written")
}

# another better way to do it
if (file.exists("mtcars.tsv")) {
  warning("File already exists!")
} else {
  write.table(mtcars,file="mtcars.tsv",quote=FALSE,sep="\t")
  message("File written")
}

# delete the file
unlink("mtcars.tsv")

# another better way to do it
if (file.exists("mtcars.tsv")) {
  warning("File already exists!")
} else {
  write.table(mtcars,file="mtcars.tsv",quote=FALSE,sep="\t")
  message("File written")
}
```
We can use Boolean operators to come of with more sophisticated conditional:
```{r}
# & = AND
if(3>2 & 5>4) {print("ok")}
if(3>2 & 5>10) {print("ok")}
# | = OR
if(3>2 | 5>10) {print("ok")}
```
# Loops
In data analysis it is common to repeat a task with different inputs. In most computer languages a `for` loop would be used. This can be done in R like this:
```{r}
for (i in c("a","b","c")) {
  print(i)
}

for (i in 1:5) {
  print(i^2)
}
```
While this works in the above example, you'll find that it is difficult to work downstream of the loop. 
```{r}
# try storing the results
x <- for (i in 1:5) {
  print(i^2)
}
x
```
For loops are also really slow and inefficient in R. Nevertheless, there is also a `while` loop.
```{r}
i=0
while(i<5) {
    print(i)
    i=i+1
}
```
# Vectorisation with sapply and lapply
The preferred solution to repeat a task on many inputs is to use vectorisation. Let's say we want to find the median of three matrices. We create a new object, either a vector or list and then run a function on each element.
```{r}
df1<-matrix(rbinom(100,50,0.8),nrow = 5)
df2<-matrix(rbinom(100,100,0.8),nrow = 5)
df3<-matrix(rbinom(100,300,0.8),nrow = 5)
dfs <- list(df1,df2,df3) 
# sapply output is always a vector
sapply(dfs,median)
# lapply output is always a list
lapply(dfs,median)
```
As you can see, vectorisation leads to shorter and more readable code.
Naturally, you can use your custom functions with `sapply` and other `apply` family functions. Here I've created a function to calculate the mean correlation in the columns of a matrix.
```{r}
meancor <- function(x) {
    mycor <- cor(x)
    mean(mycor)
}
sapply(dfs,meancor)
```
# apply
Sometimes you have a large data table and you want to run the same analysis on each row or column. To do this, use `apply`. The general syntax is:
```{r,eval=FALSE}
apply(x,MARGIN,FUN)
```
Where `x` is the object (df or mx) to work on. MARGIN is either 1 for rows and 2 for columns. FUN is the function to ron on the data.
```{r}
# row medians
apply(df1,1,median)
# column medians
apply(df1,2,median)
```
# replicate
This is a special case of sapply which repeats a task a set number of times, which could be useful for simulations (and assessment 3). In the example below, we create a function which simulates a coin toss. Tails=0 and heads=1. Calculate the mean of n=10 tosses. Is the result consistent?
```{r}
cointoss<-function(){
    sample(c(0,1),1,replace=TRUE)
}
mean(replicate(10,cointoss()))
mean(replicate(100,cointoss()))
mean(replicate(1000,cointoss()))
mean(replicate(10000,cointoss()))
```

# Test your understanding

### Question 1
Create a function that calculates the cubed root for a number

### Question 2
Use a for loop to calculate the cubed root for the numbers 44,77,99,123

### Question 3
Use `sapply` to calculate the cubed root for the numbers 44,77,99,123

### Question 4
Use `if` to test whether object `x` exists and run `head` if it does exist.

### Question 5 
Use `apply` to get the maximum value in each column of `df1`.
`df1<-matrix(rbinom(100,50,0.8),nrow = 5)`

# Session
```{r}
sessionInfo()
```