---
title: "Bioinformatics prac 2: Working with data tables"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
theme: cosmo
---

Source: https://github.com/markziemann/SLE712_files/blob/master/BioinfoPrac2.Rmd

## Introduction

In this session we will be building directly on top of what we learned last session.
Last session we introduced simple data structures in R, which are individual numbers, integers, 
strings, and vectors of these.
In this session we will be working with data tables, which are the most common type of data you will
be using.
The purpose of this session will be to become so familiar with R data tables, that you will no 
longer need to use spreadsheets in your research.

The benefit of using R over excel are many:

1. The code is totally reproducible. 

2. The code documents each and every step that you have done.

3. Errors can be picked up more easily in code than in large spreadsheets.

4. Reanalysis is quicker.

5. We can use the source control to manage different versions of the analysis (more on this next week).

6. R has many more functions than Excel.

## Data table types

When I say data table, I mean R objects that have two dimensions, similar to rows and columns in a 
spreadsheet. 
There are two main types of R data table objects, matrix and data frame.
These are distinct objects in that matrices are limited to numerical values while data frames are more
flexible.
The columns of the data frame can have different types of data like numerical, factor, string, etc.
Because of these differences, some commands that work for matrices might not work for data frames, 
so it is important to know how to find this information out.

In the below example we look at `freeny.x` which is a matrix, and `mtcars` which is a data frame.
Again, using the `str()` command will tell us the structure of the data.

```{r,datatables1}

head(freeny.x)
str(freeny.x)

head(mtcars)
str(mtcars)

```

As an example of commands that can be used for matrices and not data frames, try `mean()` for `freeny.x` and `mtcars`.

## Basic commands that can be used for data frames and matrices

Find out the number of rows, columns, and simple operations on those rows and columns.

```{r,datatables_commands}

# number of columns
ncol(mtcars)

# number of rows
nrow(mtcars)

# dimensions
dim(mtcars)

# analysing rows and columns
colMeans(freeny.x)
colSums(freeny.x)
rowMeans(freeny.x)
rowSums(freeny.x)

```


You can also transpose a matrix or data frame.
But be careful, transposing a data frame will automatically convert it to a matrix which could cause downstream errors.

```{r,transpose1}

freeny_flip <- t(freeny.x)
head(freeny_flip)

mtcars_flip <- t(mtcars)
head(mtcars_flip)
str(mtcars_flip)

mtcars_flip <- as.data.frame(mtcars_flip)
str(mtcars_flip)

```

## Subsetting a data frame

One of the most common tasks in data analysis is to perform filtering.
Last prac, we found out how to do this with vectors using the square bracket notation. Eg:`x[3]` will retrieve the 3rd element of `x`.
Square brackets can also be used for two dimensional objects, but we need to provide two indexes.
The syntax is `df[rows,cols]`.

```{r,datatables2}

# get rows 1-10 of column 2
freeny.x[1:10,2]

# get rows 1-6 of columns 1-3
freeny.x[1:6,1:3]

# get rows 1-6 of all columns
freeny.x[1:6,]

# get all rows for columns 1 and 2 
freeny.x[,1:2]

```

Now we need to see what happens when we subset just one column or row.
You can see the default behaviour is to convert the data from matrix format to a vector.
We can modify this using drop=FALSE to keep it in matrix format.

```{r,datatables3}

# get all rows of column 1
freeny.x[,1]

# to prevent conversion to vector, use drop=FALSE
freeny.x[,1,drop=FALSE]

# the same concept for rows
freeny.x[1,]

# the same concept for rows
freeny.x[1,,drop=FALSE]

```

Square brackets also works for data frames.

```{r,datatables4}

mtcars[1:10,1:6]

mtcars[,1]

mtcars[,1,drop=FALSE]

```

Data frames also have more options around subsetting columns.
For example we can subset based on the name of the column or row.

```{r,datatables5}

mtcars[,"cyl"]

mtcars[,c("mpg","wt")]

mtcars["Camaro Z28",c("mpg","wt")]

```

Data frames columns can also be subsetted using the `$` notation.
The syntax is `df$col`.

```{r,datatables6}

mtcars$mpg

```

This type of notation can even be used to create new columns.
In this example below, we are converting the miles per gallon value to liters per 100km unit.
We are also rounding this value to three significant figures.

```{r,datatables7}

mtcars$lper100km <- 235.215 / mtcars$mpg

mtcars[,c(1,ncol(mtcars))]

mtcars$lper100km <- signif(235.215 / mtcars$mpg ,3)

mtcars[,c(1,ncol(mtcars))]

```


You may also want to subset a data frame based on the values.
Let's say you want a car with fuel consumption less than 10 L/100km.
Let's do it the hard way first.

```{r, subsetting1}

mtcars$lper100km 
mtcars$lper100km < 10
which(mtcars$lper100km < 10)
mtcars[which(mtcars$lper100km < 10),]

```

You can see that this is quite complicated.
There is an easier way using `subset()`.
Subset is also perfect for filtering based on more than one criteria using the `&` and '|' operators.

```{r, subsetting2}
mtcars
subset(mtcars,lper100km < 10)

# you want an economical AND quick car
subset(mtcars,lper100km < 10 & qsec < 18)

# you want an economical OR quick car
subset(mtcars,lper100km < 10 | qsec < 18)

```

Subset also works for strings and factors.
To look at this we need to look at the `iris` dataset

```{r, subsetting3}

head(iris)
str(iris)

setosa <- subset(iris,Species == "setosa")
head(setosa)

```

## Row and columns names

You can use the `colnames` and `rownames` to get the row or column names and even mofidy them.

```{r,rowcolnames1}

colnames(mtcars)

rownames(mtcars)

colnames(mtcars) <- c("miles per gallon",
              "number of cylinders",
              "displacement in cubic inches",
              "gross horsepower",
              "rear axle ratio",
              "weight (pounds/1000)",
              "quarter mile time in seconds",
              "V or straight cylinder configuration",
              "transmission type: auto (0) or manual (1)",
              "number of forward gears",
              "number of carburetors",
              "litres per 100km")

head(mtcars)

colnames(mtcars)[1] <- "miles per US gallon"

head(mtcars)

```

If you have a whitespace in the column or row name, it might cause problems later on with subsetting. 
In that case the column name needs to be wrapped in backticks like this.

```{r, subsetting4}

economical_cars <- subset(mtcars,`litres per 100km` < 10)

economical_cars

```

It is also useful to be able to subset a data frame based on the row names.
Let's get all the Mercedes models.
To do this, we need to introduce the `grep()` command which matches strings.

```{r,rowcolnames2}

# let's look again at the car names
rownames(mtcars)

# lets filter all the ones with "Merc in the name"
grep("Merc",rownames(mtcars))

# now lets extract out all those rows
mercs <- mtcars[grep("Merc",rownames(mtcars)),]

mercs

```

## Sorting data tables

We are going to sort our subset of economical cars by their speed based on their quarter mile time.
To do this, we need to use the `order()` command together with the square brackets.
`order()` only returns the index of the values, it doesn't actually do the sorting.
Note that `order()` default behaviour is to bring the smaller values to the top.
That can be reversed by putting a `-` before the vector being ordered.

```{r,sorting1}

economical_cars

order(economical_cars$`quarter mile time in seconds`)

sorted <- economical_cars[order(economical_cars$`quarter mile time in seconds`),]

sorted[,c(7,ncol(sorted))]

reverse_sorted <- economical_cars[order(-economical_cars$`quarter mile time in seconds`),]

reverse_sorted[,c(7,ncol(reverse_sorted))]

```

## Creating data frames and matrices

First we will create a data frame for some people who completed a survey about their height and weight.
You should always run `str()` to check that the resulting dataframe has the intended structure.
You may need to include `stringsAsFactors=FALSE` to protect character strings being converted to factors.

```{r,create_df1}

pnames <- c("Jill", "Matt", "Sam", "Amy", "Bob", "Raj")

pnames

pgender <- as.factor(c("F", "M", "F", "F", "M", "M"))

pgender

pheight <- c(164, 186, 170, 175, 178, 191)

pheight

pweight <- c(54.1, 90.3, 64.8, 66.7, 80.4, 86.9)

pweight

df <- data.frame(pnames,pgender,pheight,pweight)

str(df)

df <- data.frame(pnames,pgender,pheight,pweight,stringsAsFactors = FALSE)

str(df)

df

```

Now we might want to make the row names the name of the person.
This makes the data tidier, but it won't work if there are more then one entry with the same name.
You can use the `NULL` to delete columns.
Deleting rows can be done with square brackets.

```{r,create_df2}

rownames(df) <- df$pnames

df

df$pnames=NULL

df

# delete row 2 and 4
df <- df[-c(2,4),]

```

Now we will convert df into a matrix.

```{r,create_mat1}

as.matrix(df)

df$pgender <- as.numeric(df$pgender)

mymat <- as.matrix(df)

mymat
str(mymat)

```

We can also convert other types of data into a matrix.
To demonstrate this, I'll create some random data with `rnorm` and convert it into a matrix.

```{r,create_mat2}

mydata <- rnorm(n = 100, mean = 10, sd = 20)

mymatrix <- matrix(data = mydata, nrow = 20, ncol = 5)

mymatrix

```

## Creating charts from data frame and matrix objects

Making charts from data frames and matrix objects is really similar to what we did last week.
Here are some examples.

```{r,plots1,fig.height=8,fig.width=8}

# histogram of cylinders
data(mtcars)
hist(mtcars$cyl,xlab="number of cylinders", main="number of cylinders in mtcars data")

# boxplot of qsec values for Toyota and Mercedes
toyota <- mtcars[grep("Toyota",rownames(mtcars)),]
merc <- mtcars[grep("Merc",rownames(mtcars)),]
boxplot(toyota$qsec, merc$qsec,
        ylab="quarter mile time in seconds",
        main="mtcars",
        names = c("Toyota","Mercedes"))

# scatterplot of petal length vs sepal length for setosa irises
# include a trend line using the lm function
setosa <- subset(iris,Species=="setosa")
head(setosa)
mylm <- lm(setosa$Sepal.Length ~ setosa$Petal.Length)
plot(setosa$Petal.Length,setosa$Sepal.Length, 
     xlab="Petal length (cm)",
     ylab="Sepal length (cm)",
     main="Setosa petal and sepal length",
     pch=19)
abline(mylm,col="red",lwd=2,lty=2)

# a pairs plot is a special type of scatterplot
pairs(mymatrix)

# lets make a line diagram of freeny revenues
# first need to normalise each column to the initial value
head(freeny.x)
freeny_norm <-t( t(freeny.x) / freeny.x[1,] )
head(freeny_norm)
plot(freeny_norm[,1], type="b", ylim=c(0.9,1.12), col="blue",
     xlab="Quarters beginning 1962 Q2",
     ylab="Change in values overtime")
  lines(freeny_norm[,2], type="b", col="red")
  lines(freeny_norm[,3], type="b", col="black")
  lines(freeny_norm[,4], type="b", col="darkgreen")
  legend("topleft", legend = colnames(freeny.x), lty=1 , col = c("blue","red", "black", "darkgreen"))
  
```


In R, sometimes we need to load a particular package in order to make a special type of chart.
In the example below we are making a heatmap, where the colour indicates the numerical value.

```{r,heatmaps,fig.height=8,fig.width=8  }

library("gplots")
heatmap.2(mymatrix,trace="none",scale="none",main="heatmap")
heatmap.2(cor(mymatrix),trace="none",scale="none", main="correlation")

```

This is also a good time to let you know that you can save R charts as files.
There are different types of file formats, but PNG and PDF are the most used types.
You will see these new files appear in your files menu.

```{r,output_plots}

pdf("myplot.pdf")
plot(1:10)
dev.off()

png("myplot.png")
plot(1:10)
dev.off()

```

## Check your understanding so far

Answer the questions which use concepts from this week and previous sessions.

1. Create a scatterplot of mtcars weight (x axis) versus mpg (y axis). 
Include x and y axis labels and a main heading.

2. Sort mtcars by weight (wt) and create a horizontal barplot of wt values so heaviest ones at shown at the top of the bar plot.
The plot should be labeled so it is clear to see which bar belong to which car.
Include an axis label and main title.

3. Create a box plot of iris petal lengths.
Each species should be a different category.
Chart needs axis labels and a main title.

We will discuss your solutions next week.

## Reading in data files

So far we have been working with datasets that are built into R, but in the real world you will be working with data files.
These could be in different formats like text (.txt), comma separated values (.csv), tab separated values (.tsv) and perhaps Excel files too (.xls and .xlsx).

We will be working with TSV file of pipetting measurements. URL: https://raw.githubusercontent.com/markziemann/SLE712_files/master/pipette_test.tsv

We will use the `read.table()` command and show you some really important options.

```{r,tsv}

URL="https://raw.githubusercontent.com/markziemann/SLE712_files/master/pipette_test.tsv"

# look closely at the structure of pip
# can you see what is wrong?
pip <- read.table(URL)
pip
str(pip)

# try again
pip <- read.table(URL,stringsAsFactors = FALSE)
pip
str(pip)

# try again
# looking better
pip <- read.table(URL,stringsAsFactors = FALSE, header=TRUE)
pip
str(pip)

# got it now
pip <- read.table(URL,stringsAsFactors = FALSE, header=TRUE, row.names=1)
pip
str(pip)

```

Now let's try a csv file containing some travel records.

```{r,readcsv}

URL="https://people.sc.fsu.edu/~jburkardt/data/csv/airtravel.csv"

trav <- read.table(URL,sep=",")
trav
str(trav)

trav <- read.table(URL,sep=",",header=TRUE)
trav
str(trav)

trav <- read.csv(URL)
trav
str(trav)

```

Now let's try a Microsoft Excel file.

```{r,readxl}

URL="https://github.com/markziemann/SLE712_files/blob/master/misc/file_example_XLS_10.xls?raw=true"
NAME="file_example_XLS_10.xls"
download.file(URL,destfile=NAME)
library("readxl")

mydata <- read_xls(NAME)
mydata
str(mydata)

```

## Save and load session and single datasets

When working in R, it is convenient to save the session with `save.image()`. 
This results in an Rdata file which contains all the data objects in your current environment.
To test that it's actually working, clear your environment with the sweep/broom icon and then load the Rdata file with `load()`.

```{r,save1}

save.image("mysession.Rdata")

rm(list=ls())

load("mysession.Rdata")

```

That is really cool, but sometimes we want to save individual objects, such as a large dataframe as Rdata files.

```{r,save2}

saveRDS(object = mymatrix , file = "mymatrix.Rds")

rm(list=ls())

x <- readRDS("mymatrix.Rds")

head(x)

```

## Check your skills

For the TSV file located here: https://raw.githubusercontent.com/markziemann/SLE712_files/master/misc/mydata.tsv

1. Read it in and show the first 6 rows of data. 

2. Calculate the column and row means.

3. Use the `cor()` command to find the correlation coefficients between the 3 data sets. Which two datasets are the most similar?

4. Make a pairs plot of the three datasets.


## Session information

For reproducibility.

```{r,sessioninfo}

sessionInfo()

```
