---
title: "R Cheatsheet"
author: "Mark Ziemann"
date: "27/04/2020"
output:
    rmarkdown::html_document:
        toc: true
theme: cosmo
---
Source: https://github.com/markziemann/SLE712_files/cheatsheet.Rmd

In this document we will be revising everything we covered in our Rstudio work in the form of a cut-down R markdown file.

## Week 1
### Arithmetic
We first covered arithmetic. Here are some examples.
```{r}
1+2
# comments can be placed after a hash
10000^2
sqrt(64)
2^3
27^(1/3) # cube root
27^1/3 # parentheses matter!
```

### Math on vectors
Math can also be done on vectors (a vector is a series of numbers). Let's first save the vector as `a`.

```{r}
a <- c(1,10,100) # use the c() to create a vector
sum(a) # calculate the sum of a vector
mean(a) # calculate the mean of a vector
median(a) # calculate the median of a vector
max(a) # show the maximum value in a vector
min(a) # show the minimum value in a vector
range(a) # show maximum and minimum value
?range # use the ? to get help on a command
```

### Understanding object types
It is important to understand object types. It is a major source of errors in R. Currently `a` is a numerical vector. Here are some commands to determine an object's type.

```{r}
str(a) # get data structure
typeof(a) # get data type (can be deceiving)
class(a) # get data class

b = "x1"
str(b)
typeof(b)
class(b)

c <- c("abc","def","ghi")
str(c)
typeof(c)
class(c)
```

### Square brackets
Subsetting vectors can be done with square brackets `[]`.

```{r}
a[2]  # get 2nd value
a[2:3] # get values in order
a[c(1,3)] # get values out of order
a[c(2,1)] # get values in reverse order
```

### Reordering vectors
Reordering vectors can be done with square brackets `[]` as well.
```{r}
order(a) # rank the values from lowest to highest
order(-a) # rank values for highest to lowest
a[order(-a)] # reorder from highest to lowest 
a <- a[order(-a)] # reorder and save as a
```

### Diagnosing object types/classes
We can find out a data type with the following commands.

```{r}
is.numeric(a) # check if numeric (value, vector or table)
is.character(b) # check if numeric (value, vector or table)
is.vector(c) # check if structure is vector or other data  type such as list or df
is.list(a) # check if structure is list
is.integer(a) # check if data is integers
```

### Coercing types
We can coerce some data types to others.

```{r}
# for example, data can be imported as strings
d <- c("1","3","5","654")  # create a new vector of strings
str(d) # confirm the structure
d <- as.numeric(d) # then convert to numeric
str(d) # confirm the structure
d <- as.integer(d) # coerce to integer
str(d) # confirm the structure
# we can give vector values names
names(d) <- c("s1","s2","s3","s4")
names(d) # show the names
str(d) # named vector looks a bit different in str
d <- as.list(d) # coerce data to a list
str(d) # confirm the structure
d # have a look at the object in the console
```

## Pause and check your understanding
At this point you will be getting familiar with some of R's data types. Can you describe clearly what the differences are between character strings, numerical values, vectors and lists? Do some reading online and take some notes if necessary.

## Two-dimensional data tables
Data frames and matrices are important data types in R.

### Converting vectors and dataframes
We can convert vectors to data frames and vice versa.

```{r}
b<-c(1.23,2.34,4.87,21.1) # create a new numerical vector
b<-as.data.frame(b) # coerce to df
str(b)
b
# set row names
rownames(b)<-c("sample1","sample2","sample3","sample4") 
b
colnames(b)="B" # set the column names
b
```

### The dollar-sign notation
The dollar sign $ allows us to address individual columns of a dataframe.
If we specify a column name that doesn't exist, we can create a new column with that name.

```{r}
b$B # extract a column of data from a df and show as a vector
b$D <- c(8.45,6.15,47.4,56.1) # create a new column of data
b
b$F<-b$B+b$D # doing math with columns in dataframes!
b
```

It's possible to do some mathematical operations on a column or whole df.
```{r}
b$D*100 # multiply a column by a number
b*100 # multiply a whole data frame by a number
b^2 # square each value of a data frame
str(b) # str makes sense most of the time
class(b) # makes sense as well
typeof(b) # sometimes deceptive
str(as.matrix(b)) # see how matrix is stored differntly internally
str(b)
```


### adding columns and rows to a dataframe
`cbind` adds new columns and `rbind` adds new rows.

```{r}
G<-c(78.03,54.8,98.1,0.2) 
b<-cbind(b,as.data.frame(G)) # bind columns together
b
sample5 <- c(12.1, 3.84, 549.1, 111.9)
b<-rbind(b,sample5,make.row.names =T) # bind rows together
b
rownames(b)[5]<-"sample5" # change a single row name
b
```

### other operations we can do on a dataframe or matrix
Here are some key commands that will help you analyze data and complete Assignment 3 as well.

```{r}
t(b) # transpose a df or matrix
nrow(b) # get number of rows
ncol(b) # get number of columns
dim(b) # get number of rows and columns
rowMeans(b) # get means of all rows
colMeans(b) # get means of all columns
rowSums(b) # get sum of all rows
colSums(b) # get sum of all columns
summary(b) # get a nice summary of each column
mean(b) # one does not simply calculate mean of a dataframe
mean(as.matrix(b)) # but it does work for a matrix
```
### Dot and line plots
Now let's explore plots
```{r}
plot(b$B) # dot plot a vector of values 
plot(b$B, type = "l") # line plot a vector of values 
plot(b$B, type = "b") # plot both lines and dots
plot(b$B, b$F) # scatter plot of two vectors
plot(b$B, b$D, pch=19) # pch=19 gives nice circles
points(b$F,b$G,col="red",pch=19) # add red points. But make sure axis limits are wide enough!!
mtext("B vs D in black ; F vs G in red") # add a sub-heading
```

### Barplot and histograms
So that's how to make a basic dot/line plot. Make sure you add axis labels with `xlab=` and `ylab=` options. Include the units of measurement and a main heading or subheading to better describe the data. Now let's look at some other basic chart types.

```{r}
barplot(b$B,names.arg=rownames(b)) # barplot works nicely if you supply the names.arg
# here is how you can add axis labels and main header
barplot(b$B,names.arg=rownames(b),ylab="Elapsed time (s)",
    main="Time trial")
# sometimes a horizontal chart looks better
barplot(b$B,names.arg=rownames(b),xlab="Elapsed time (s)",
    main="Time trial",horiz=T, las=2) 
# here's how to make a histogram
hist(as.matrix(b)) 
# alter the number of histogram bins
hist(as.matrix(b),10)
```

### Histogram and boxplot with simulated data
The above histogram example uses a very small dataset. I think it would be better if we could use a bigger one. So lets create some new data with `rnorm`. This is a command to sample data from a distribution. Then we can get a nice looking histogram and introduce boxplot. Can you see the difference in the distributions in object `d` and `e`?

```{r}
d <- rnorm(n = 100,mean = 10,sd = 20) # sample 100 values from a distribution with mean=10 and SD=20
d
e <- rnorm(n = 100,mean = 10,sd = 50) 
hist(d)
hist(e)
boxplot(d,e) 
```

# Export the PDF file
If you want to download the chart, you can use the export button in the lower right Rstudio window. A better alternative, especially for multiple charts is to create a multi-page PDF then download this file.

```{r}
pdf("myplots.pdf") # create a new, empty PDF
hist(d) # then add charts one by one
hist(e) # you can add lots and lots
boxplot(d,e) # and lots
dev.off() # when finished, you need to run dev.off() otherwise the PDF will appear as corrupted.
```

Check whether the file has been created in your current working directory and click on it to open in the web browser and find the button to save it.

## Week 2
### Factors
Let's work with factors. Factors are an efficient way to store categorical data.

```{r}
g<-as.factor(c("M","M","F","F","M")) # in this command we are coercing strings to factors
str(g) # see how factors are represented in the object g
# We can specify the levels like this
f<-factor(c("early","late","ontime","early","late","ontime","early"), 
          levels = c("early", "ontime", "late")) 
f
str(f)
# Some factors are "ordinal" in that they have values that relate to each other
# such as high medium and low. we can specify this
f<-factor(c("early","late","ontime","early","late","ontime","early"), 
          levels = c("early", "ontime", "late"), ordered =TRUE) 
f
str(f)
as.character(f) # coerse factor to character
as.numeric(f) # coerce factor to numeric

```

### Subsetting data frames part 1
We can subset a column from a data frame with the `df$col` notation. With the `[rows,cols]` square brackets we can select both rows and columns.
```{r}
data(mtcars) # make sure that mtcars is available
mtcars # show mtcars in the console
mtcars$mpg # extract one column with $ notation
mtcars[,"mpg"] # it can also be done with [] notation by specifying the col name
mtcars[,1] # or specifying the location. Here we extract  the 1st col
mtcars[,1:3] # range of columns
mtcars[,c(3,1,8)] # columns in different order
mtcars[1:4,] # rows as well
mtcars[1:4,1:6] # rows and columns
```
### Subsetting data frames part 2
The above is useful, but most subsetting jobs are based on the values inside the data frame. 

```{r}
mtcars$mpg>20 # ask the question which mpg values are >20
mtcars[,"mpg"]>20 # same thing
mtcars[,1]>20 # same thing again
which(mtcars[,1]>20) # Which command extracts all the TRUE values
mtcars[which(mtcars[,1]>20),] # use which inside the square brackets to subset cars with mpg>20
subset(mtcars, mpg>20) # a simpler way to do it.
subset(mtcars, mpg>20 & cyl>5) # multiple conditions with AND
subset(mtcars, mpg>20 | cyl>5) # multiple conditions with OR 
```

While `subset` command is easier, you still need to know the `[]` way of doing it because that is more versatile and used in the next section.

### Reordering a data frame
It is a very common procedure to reorder a dataframe based on the values. Note that it is a 2 step process. 1st to rank the values and 2nd to extract rows/cols in that order. This is likely to be on the exam. 

```{r}
order(mtcars$mpg) # rank the values in a column (vector)
mtcars[order(mtcars$mpg),] # stick the order command in the square brackets to extract the rows in order.
mtcars[order(-mtcars$mpg),] # reverse the order
mtcars[order(rownames(mtcars)),] # reordering based on the alphabetical order of the rownames
```

### Linear regression example
Linear regression is one of those fundamental, essential skills in R. Let's take a look at the correlation between number of cylinders and mpg (or 1/mpg).

```{r}
mtcars[,1:2] # extract 1st and 2nd column
cor(mtcars[,1:2]) # simple correlation
pairs(mtcars[,1:2]) # pairs plot can be useful to see relationships
plot(mtcars$cyl,mtcars$mpg) # scatterplot
plot(mtcars$cyl,1/mtcars$mpg) # 1/mpg gives the fuel economy
plot(mtcars$cyl,1/mtcars$mpg,xlab="cylinders",ylab="gallons per mile",pch=19) 
```

In any chart it is really important to give meaningful x and y labels.

```{r}
mylm<-lm(1/mpg ~ cyl, data=mtcars) # calculate the linear model
str(mylm) # the output is a big list object
summary(mylm) # summary helps to make sense of it
summary(mylm)$r.squared # we can extract values with summary
signif(summary(mylm)$r.squared,3)  # signif helps to limit the significant figures for presentation
plot(mtcars$cyl, 1/mtcars$mpg, xlab="cylinders", 
     ylab="gallons per mile", pch=19) 
abline( mylm , lty=2,lwd=2,col="red") # make trend line
# now we can extract some other values from the lm object to make a nice plot
INT=as.numeric(signif(mylm$coefficients[1],3))
SLOPE=as.numeric(signif(mylm$coefficients[2],3))
MYR2=signif(summary(mylm)$r.squared,3)
MYTEXT=paste("Intercept:",INT, " Slope:",SLOPE, " R^2:",MYR2)
mtext(MYTEXT)
```

### Reading data from a file
Reading in data from a file improperly is a major source of errors. For example if R doesn't recognise column headers, then it can lead to the data being recognised as strings instead of numerical values. In the worst cases, the strings are represented as factors; then when maths operations are done, the user gets results which are totally wrong. Let's have a look at the following typical example.

```{r}
# you will need to download the file first
download.file(url = "https://raw.githubusercontent.com/markziemann/SLE712_files/master/bioinfo_asst3_part1_files/mydata.tsv", destfile = "mydata.tsv")
# read.table a the command to read a table of data
head(read.table("mydata.tsv"))
x<-read.table("mydata.tsv") # typical way to import data, but be careful about how its imported
typeof(x)
str(x) # check the data. Is it saved as numeric??
x <- read.table("mydata.tsv",header=TRUE ) # set header=TRUE if you have headers on your data
str(x)
x <- read.table("mydata.tsv", header=TRUE, row.names=1) # set row.names=1 if the 1st col is the row name
head(x)
str(x)  # check that it's numeric
# here is another correct way to import especially if you have gene names (strings)
x<-read.table("mydata.tsv",header=T,
  stringsAsFactors=FALSE,sep="\t")
str(x)
# another correct way to import the data
x<-read.csv("mydata.tsv",sep="\t", stringsAsFactors=FALSE)
str(x)
```

### Saving R sessions
This can help you save time especially if you have been running a large data analysis, you can save the results for future reference so you don't need to execute it again.

```{r}
# save a session
save.image("mydata.RData")
# now in the menu bar select “quit session”
# then start new session
# see if x is saved
x
# load session
load("mydata.RData")#    see if x is saved x
```

### Session Info
```{r}
Sys.Date()
sessionInfo()
```
