---
title: "Home exercise: check your R understanding"
author: "Mark Ziemann"
date: "20/04/2020"
output: html_document
theme: cosmo
---

Source: https://github.com/markziemann/SLE712_files/blob/master/home_exercise.Rmd

### Question 1

> Do you know the difference between Rstudio and R?

R is the computing language, whereas Rstudio is the interface (integrated development environment). Rstudio has several features which make coding in R easier than just typing commands into the console. It has a text editor with syntax highlighting, global environment contents, file navigator, git integration, plot window and other features.

### Question 2
> Can you make a histogram or box plot of mpg values from the mtcars dataset?

The `boxplot` and `hist` commands can be used to create boxplotsand histograms. The argument `mtcars$mpg` should be used.
```{r,question2,echo=TRUE}
boxplot(mtcars$mpg)
hist(mtcars$mpg)
```

### Question 3
> Can you subset a dataframe like mtcars to identify the cars with mpg greater than 25?

There are two common ways to do this; with `subset` and with the square brackets `[]`.
```{r,question3,echo=TRUE}
# with subset command
subset(mtcars,mpg>25)
# or with square brackets
mtcars[mtcars$mpg>25,]
```

### Question 4
> Can you re-order a dataframe like mtcars by mpg values and save it as a different object? 

This is tricky and I normally take a 2 step approach with `order` inside square brackets `[]`.
```{R,question4,echo=TRUE}
# order(mtcars$mpg)
# mtcars[rows,cols]
# use <- to save to a different object
mtcars2 <- mtcars[order(mtcars$mpg),]
```

### Question 5
> How can you properly import a csv and tsv file? 

Importing data in csv and tsv files needs to be done carefully. A command such as read.table can be used, but the options need to be set correctly. If the data is in tab separated format, then `read.table` will recognise it. Sometimes setting the separator to "\t" is necessary. If it is a csv file, then `read.csv` would be the best function to read it. if the data file has headings (column names), then `headers = TRUE` needs to be set. If the first row contains row names, then the `row.names=1` option should be set. if the table contains data in strings then we will need to set the `stringsAsFactors=FALSE` option because the default behavior is to convert non numerical data such as strings to factors. Finally, commands `head` and `str` should be used to confirm that the data have been imported properly.

```{r,question5,echo=TRUE}
# first write out a table to a file in tsv format
write.table(mtcars,file="mtcars.tsv",quote=FALSE,row.names=TRUE,sep="\t")
# read in the tsv file 
x<- read.table("mtcars.tsv",header=TRUE,row.names=1,sep="\t")
head(x)
str(x)
# first write out a table to a file in csv format
write.table(mtcars,file="mtcars.tsv",quote=FALSE,row.names=TRUE,sep=",")
# read in the tsv file 
x<-read.table("mtcars.tsv",header=TRUE,row.names=1,sep=",")
head(x)
str(x)
```
### Question 6
> Can you save an Rsession to a file  and then load it again later?

This can be done fairly easily using the commands shown below:
```{r,question6,echo=TRUE}
save.image("mysession.RData")
load("mysession.RData")
```

### Some information about the R version for reproducibility
```{r,session,echo=TRUE}
sessionInfo()
```